Structuring College Access: The Market Segment Model and College Board Geomarkets

### 2025-04-15 text/back_end_KS.qmd
smooth rq2 results chi

To examine how the socioeconomic and racial composition of included versus excluded prospects varies when student list purchases filter on particular Geomarkets, we analyze six orders placed by a research university targeting prospects across the entire state of Illinois. Two of these orders applied the same filters: relatively lower SAT score thresholds ranging from 1020 to 1150, GPA ranging from B- to A+, and high school class targeting students from the 2019-2020 and 2020-2021 graduating classes, respectively. The remaining four orders used similar GPA and high school class filters but differed in SAT thresholds- two targeted students with middle-range scores (1160 to 1300) and two targeted those with relatively high scores (1310 to 1600). Because all six orders included an Illinois state filter, we can utilize prospects from the Chicago metropolitan area within the resulting student lists to examine Geomarket contributions by simulating the application of a Geomarket filter across low, mid, and high range score threshold orders.

@@

In contrast, these representational patterns are reversed for Black and Hispanic prospects in @fig-chicago-rq2-lowSAT-race. Among the more than 2,000 Black students included in these orders, the City of Chicago and South and Southwest Suburbs contributed disproportionately higher shares - 48% and 34%, respectively- compared to their overall contributions of 20% and 26% of all prospects whereas all other Geomarkets contributed substantially smaller proportions. The City of Chicago Geomarket also contributed larger shares of the nearly 5,000 Hispanic students (35% versus 20% for all prospects), while all other Geomarkets contributed comprable or relatively smaller proportions than their overall representation among all included prospects.

@fig-chicago-rq2-midSAT-race and @fig-chicago-rq2-highSAT-race similarly depict Geomarket contributions for pools of prospects resulting from orders targeting middle-range (1160-1300) and high-range (1310-1600) SAT scores. Both figures demonstrate how representational Geomarket patterns remain relatively consistent SAT scores increase. For instance, @fig-chicago-rq2-midSAT-race shows that the Chain of Lakes, North Shore, and Western Suburbs Geomarkets contributed similar proportions of White and Asian students, yet smaller proportions of Black and Hispanic students, relative to their overall representation among all included prospects. The Northwest Suburbs and Evanston and Skokie similarly contributed a disproportionately higher share of Asian prospects, but lower shares of Black and Hispanic students. Conversely, the South and Southwest Suburbs Geomarket contributed a substantially larger proportion of Black students, while the City of Chicago yielded larger proportions of both Black and Hispanic students- again, relative to their overall representation among all included prospects targeted at middle range SAT scores. Despite declines in the overall number of Black (n=240) and Hispanic (n=950) prospects included in orders filtering for high-ranges of SAT scores (1310-1600), @fig-chicago-rq2-highSAT-race shows overall Geomarket contribution patterns persist.

We also examine the contributions of each Geomarket to the pool of prospects based on first-generation college student status. Similar to race/ethnicity figures above, @fig-chicago-rq2-lowSAT-ses presents the Geomarket distribution for the 16,060 total prospects (whose first-generations status is known) included in low-range SAT score orders (1020-1150). Among the 3,479 prospects whose parents did not attend college, the City of Chicago stands out as the only Geomarket that contributing a disproportionately larger share of first-generation students - 34% in comparison to 20% for all prospects in this SAT range. In contrast, all other Geomarkets contributed disproportionately smaller shares, ranging from a one-percentage-point difference (Chain of Lakes, Northwest Suburbs, North Shore) to a five-percentage-point difference (Western Suburbs). For prospects whose parents attended some college but did not complete their degrees, @fig-chicago-rq2-lowSAT-ses indicates both City of Chicago and South and Southwest Suburbs Geomarkets contribute a disproportionately larger share of first-generation students. These representational patterns are reversed among prospects that are not first-generationcollege students. While all other Geomarkets contributed either comparable (Northwest Suburbs, South and Southwest Suburbs) or disproportionately larger shares (Chain of Lakes, North Sore, Evanston and Skokie, Western Suburbs), the City of Chicago Geomarket contributed a substantially smaller share of not-first generation prospects (13%) relative to all prospects (20%) in this SAT range.

@fig-chicago-rq2-midSAT-ses and @fig-chicago-rq2-highSAT-ses demonstrate Geomarket contribution patterns by first-generation college student status become increasingly pronounced in orders targeting middle-range (1160-1300) and high-range (1310-1600) SAT scores, respectively. At the middle SAT score range, the City of Chicago Geomarket contributes a substantially larger share of first-generation students whose parents did not attend college—38%, compared to 17% of all prospects in this range, representing a 21-percentage-point disparity. This disproportionality increases to a 25-percentage-point difference at the highest SAT score thresholds, where the City of Chicago accounts for 40% of such first-generation students, despite representing only 15% of all prospects in this score range.

Finally, we examine Geomarket contributions to prospect pools by both first-generation college student status and race/ethnicity. @fig-chicago-rq2-lowSAT-combo presents the Geomarket distribution of the 15,183 total prospects (whose first-generation status and race/ethnicity is known) included in low-range SAT (120-1150) orders. For instance, among all White prospects (n=7,605) whose contact information was purchased in this SAT score range, approximately 11% were first-generation college students whose parents did not attend college, 22% had parents who attended some college but did not complete their degree, and 67% were not first-generation college students. @fig-chicago-rq2-lowSAT-combo shows the City of Chicago and South and Southwest Suburbs Geomarkets disproportionately contribute larger shares of White first-generation college students who parents did not attend college (22% and 15%, respectively) and whose parents attended some college but never completed degrees (29% and 28%, respectively). In contrast, Geomarkets such as Chain of Lakes (72%), North Shore (88%), Evanston and Skokie (73%) contribute a disproportionately larger share of White prospects who are not first-generation college students.

@fig-chicago-rq2-lowSAT-combo indicates these Geomarket patterns persist across other racial/ethnic categories. The City of Chicago Geomarket consistently contributed a disproportionately large share of first-generation college students whose parents did not attend college across all racial/ethnic categories. However, the magnitude of this disproportionality varied- from a 3-percentage-point difference for Black prospects to a 21 percentage-point difference for Asian prospects. Among Hispanic students, contributions to the first-generation prospect pool were notably balanced across Geomarkets, with the exception of North Shore and Evanston and Skokie Geomarkets, which contributed substantially lower shares. Although Asian and Black prospects who are not first-generation college students make up a relatively smaller portion of the overall pool, @fig-chicago-rq2-lowSAT-combo suggests that such prospects are primarily contributed by the Chain of Lakes, North Shore, and Evanston and Skokie Geomarkets.

Geomarket contribution patterns become more pronounced as SAT score thresholds increase. For instance, @fig-chicago-rq2-midSAT-combo presents prospect pools disaggregated by both first-generation college student status and race/ethnicity within the middle-range SAT score (1160–1300) orders. The City of Chicago Geomarket still contributes a disproportionately larger share of first -generation college students - both those whose parents did not attend college and those whose parents attended some college but did not complete a degree - across most racial and ethnic groups. However, the magnitude of this disproportionately for City of Chicago becomes even greater for Asian students, with a nearly 30-percentage-point difference collectively for all first-generation college students (35% no college and 22% some college) relative to their overall Geomarket representation (28%). Similar to patterns observed at lower SAT score thresholds, the Geomarket contributions to first-generation Hispanic college students remain relatively balanced. All Geomarkets, with the exception of North Shore (22%) and Evanston and Skokie (32%), contributed more than 50% first-generation college students across both parental education categories to the overall prospect pool.

@fig-chicago-rq2-highSAT-combo shows Geomarket patterns become even more pronounced at the highest SAT score range (1310-1600), largely driven by the relatively fewer number of first-generation college students included within these orders. Again, the City of Chicago Geomarket disproportionately contributes a disproportionate share of first-generation Asian college students (43% across both parental education categories) and first-generation Hispanic college students (52% across both parental education categories) in comparison to the overall distribution of first-generation college students within each racial/ethnic category (13% for Asian and 40% for Hispanic, respectively). Conversely, nearly all other Geomarkets contribute a disproportionately larger share of non-first generation college students across all racial/ethnic categories.

>>>>

To examine how the socioeconomic and racial composition of included versus excluded prospects varies when student list purchases filter on particular Geomarkets, we analyze six orders placed by a research university targeting prospects across the entire state of Illinois. Two of these orders applied the same filters: SAT score thresholds ranging from 1020 to 1150, GPA ranging from B- to A+, and high school class targeting students from the 2019-2020 and 2020-2021 graduating classes, respectively. The remaining four orders used similar GPA and high school class filters but differed in SAT thresholds- two targeted students with relatively middle-range scores (1160 to 1300) and two targeted those with relatively high scores (1310 to 1600). Because all six orders included an Illinois state filter, we can utilize prospects from the Chicago metropolitan area within the resulting student lists to examine Geomarket contributions by simulating the application of a Geomarket filter across low-, middle-, and high-range SAT threshold orders.

@@

In contrast, these representational patterns are reversed for Black and Hispanic prospects in @fig-chicago-rq2-lowSAT-race. Among the more than 2,000 Black students included in these orders, the City of Chicago and South and Southwest Suburbs contributed disproportionately higher shares - 48% and 34%, respectively- compared to their overall contributions of 20% and 26% of all included prospects, whereas all other Geomarkets contributed substantially smaller proportions of Black prospects. The City of Chicago Geomarket also contributed larger shares of the nearly 5,000 Hispanic students (35% versus 20% for all prospects), while all other Geomarkets contributed comprable or relatively smaller proportions of Hispanic prospects than their overall representation among all included prospects.

@fig-chicago-rq2-midSAT-race and @fig-chicago-rq2-highSAT-race similarly depict Geomarket contributions for pools of prospects resulting from orders targeting middle-range (1160-1300) and high-range (1310-1600) SAT scores. Both figures demonstrate how representational Geomarket patterns across race/ethnicity remain relatively consistent even as SAT scores increase. For instance, @fig-chicago-rq2-midSAT-race shows that the Chain of Lakes, North Shore, and Western Suburbs Geomarkets contributed similar proportions of White and Asian students, yet smaller proportions of Black and Hispanic students, relative to their overall representation among all included prospects. The Northwest Suburbs and Evanston and Skokie similarly contributed a disproportionately higher share of Asian prospects, but lower shares of Black and Hispanic students. Conversely, the South and Southwest Suburbs Geomarket contributed a substantially larger proportion of Black students, while the City of Chicago yielded larger proportions of both Black and Hispanic students- again, relative to their overall representation among all included prospects targeted at middle range SAT scores. Despite declines in the overall number of Black (n=240) and Hispanic (n=950) prospects included in orders filtering for high-ranges of SAT scores (1310-1600), @fig-chicago-rq2-highSAT-race shows overall Geomarket contribution patterns persist.

We also examine the contributions of each Geomarket to the pool of prospects based on first-generation college student status. Similar to race/ethnicity figures above, @fig-chicago-rq2-lowSAT-ses presents the Geomarket distribution for the 16,060 total prospects (whose first-generations status is known) included in low-range SAT score orders (1020-1150). Among the 3,479 total prospects whose parents did not attend college, the City of Chicago stands out as the only Geomarket contributing a disproportionately larger share of first-generation college students - 34% in comparison to 20% for all included prospects in this SAT range. In contrast, all other Geomarkets contributed disproportionately smaller shares of first-generation college students relative to all included prospects, with such disparity ranging from a one-percentage-point difference (Chain of Lakes, Northwest Suburbs, North Shore) to a five-percentage-point difference (Western Suburbs). For prospects whose parents attended some college but did not complete their degrees, @fig-chicago-rq2-lowSAT-ses indicates both City of Chicago and South and Southwest Suburbs Geomarkets contribute a disproportionately larger share of first-generation students. These representational patterns are reversed among prospects that are not first-generation college students. The City of Chicago Geomarket contributed a substantially smaller share of not-first generation prospects (13%) relative to all prospects (20%) in this SAT range, whereas all other Geomarkets contributed either comparable (Northwest Suburbs, South and Southwest Suburbs) or disproportionately larger shares (Chain of Lakes, North Sore, Evanston and Skokie, Western Suburbs).

@fig-chicago-rq2-midSAT-ses and @fig-chicago-rq2-highSAT-ses demonstrate Geomarket contribution patterns by first-generation college student status become increasingly pronounced in orders targeting middle-range (1160-1300) and high-range (1310-1600) SAT scores, respectively. At the middle SAT score range in @fig-chicago-rq2-midSAT-ses, the City of Chicago Geomarket contributes a substantially larger share of first-generation students whose parents did not attend college—38%, compared to 17% of all included prospects in this range, representing a 21-percentage-point disparity. This disproportionality increases to a 25-percentage-point difference at the highest SAT score thresholds in @fig-chicago-rq2-highSAT-ses, where the City of Chicago accounts for 40% of such first-generation students, despite representing only 15% of all included prospects in this score range.

Finally, we examine Geomarket contributions to prospect pools by both first-generation college student status and race/ethnicity in the Chicago metropolitan area. @fig-chicago-rq2-lowSAT-combo presents the Geomarket distribution of the 15,183 total prospects (whose first-generation status and race/ethnicity is known) included in low-range SAT (120-1150) orders. For instance, among all White prospects (n=7,605) whose contact information was purchased in this SAT score range, approximately 11% were first-generation college students whose parents did not attend college, 22% were first-generation college students whose parents attended some college but did not complete their degree, and 67% were not first-generation college students. @fig-chicago-rq2-lowSAT-combo shows the City of Chicago and South and Southwest Suburbs Geomarkets disproportionately contribute larger shares of White first-generation college students who parents did not attend college (22% and 15%, respectively) and whose parents attended some college but never completed degrees (29% and 28%, respectively). In contrast, Geomarkets such as Chain of Lakes (72%), North Shore (88%), Evanston and Skokie (73%) contribute a disproportionately larger share of White prospects who are not first-generation college students.

@fig-chicago-rq2-lowSAT-combo indicates these Geomarket patterns persist across other racial/ethnic categories. The City of Chicago Geomarket consistently contributed a disproportionately large share of first-generation college students whose parents did not attend college across all racial/ethnic categories. However, the magnitude of this disproportionality varied- from a 3-percentage-point difference for Black prospects to a 21 percentage-point difference for Asian prospects. Among Hispanic students, contributions to the first-generation prospect pool were notably balanced across Geomarkets, with the exception of North Shore and Evanston and Skokie Geomarkets, which contributed substantially lower shares of first-generation Hispanic students. Although Asian and Black prospects who are not first-generation college students make up a relatively smaller portion of the overall pool, @fig-chicago-rq2-lowSAT-combo suggests that such prospects are primarily contributed by the Chain of Lakes, North Shore, and Evanston and Skokie Geomarkets.

Geomarket contribution patterns to prospect pools by both first-generation college student status and race/ethnicity become more pronounced as SAT score thresholds increase. For instance, @fig-chicago-rq2-midSAT-combo presents prospect pools disaggregated by both first-generation college student status and race/ethnicity within the middle-range SAT score (1160–1300) orders. The City of Chicago Geomarket still contributes a disproportionately larger share of first -generation college students - both those whose parents did not attend college and those whose parents attended some college but did not complete a degree - across most racial and ethnic groups. However, the magnitude of this disproportionately for City of Chicago becomes even greater for Asian students, with a nearly 30-percentage-point difference collectively for all first-generation college students (35% no college and 22% some college) relative to their overall Geomarket representation (28%). Similar to patterns observed at lower SAT score thresholds, the Geomarket contributions to first-generation Hispanic college students remain relatively balanced. All Geomarkets, with the exception of North Shore (22%) and Evanston and Skokie (32%), contributed more than 50% first-generation Hispanic college students across both parental education categories to the overall prospect pool.

@fig-chicago-rq2-highSAT-combo shows the magnitude of Geomarket contribution disparities increases at the highest SAT score range (1310-1600), partly driven by the relatively fewer number of first-generation college students included within these orders. Again, the City of Chicago Geomarket disproportionately contributes a disproportionate share of first-generation Asian college students (43% across both parental education categories) and first-generation Hispanic college students (52% across both parental education categories) in comparison to the overall distribution of first-generation college students within each racial/ethnic category (13% for Asian and 40% for Hispanic, respectively). Conversely, nearly all other Geomarkets contribute a disproportionately larger share of non-first generation college students across all racial/ethnic categories.
